---
title: "Buffalo Land Cover Mapping"
author: Ying Lu
subtitle: Project for GEO 503
---

# Introduction
In order to trace the change and improve a city development, it is important for us to map the land use for cities. However, the update of land use images published by government is relatively slow for monitoring some small changes. The public of Landsat imagery offers us an opportunity to map land cover of cities using remote sensing methods. Nevertheless, the spatial characteristic for each city is not the same, the methods suitable for every city should not be the same. In order to get the most applicable classification method for Buffalo with Landsat imagery, three commonly used classification methods -Random Forest, Support Vector Machine and Maximum likelihood- were trained and tested in Buffalo with Landsat 8 images. Random Forest performed relatively better than other two. Therefore, the land use images in 2013, 2015, 2017 were also created with the trained Random Forest model. According to the result, impervious surface covers a great part of Buffalo, which is 59.34% in 2018. The temporal land cover maps of Buffalo show that Buffalo does not change a lot in the past 5 years. This study offers a good reference for land cover map for Buffalo city and give a general information of the change of the city. 


# Materials and methods

I used a Landsat 8 image in Buffalo in Augest 23, 2018 to map the landcover of Buffalo. The landcover of Buffalo was classified into four classes: vegetation, impervious surface, soil and water. There were three main steps for this study.

1. The training and testing data were collected in the entire extent of the Landsat 8 image in Buffalo in Augest 23, 2018. ROI method in ENVI was used. Data was exported as `.shp` file. 
2. Random Forest, Support Vector Machine and Maximum Likelihood were trained and tested using the entire Landsat image. According to the confusing matrix, Random Forest method was relatively better for Buffalo. 
3. The Landsat 8 image in 2013,2015,2017 with the low cloud coverage were downloaded from USGS explorer to map the land cover in Buffalo in those dates. A croase change of Buffalo were analyzed from those maps.

The packages required in this study was listed below (some packages may need to be installed):

```{r, message=F, warning=F}
library(getSpatialData)
library(sf)
library(maps)
library(rasterVis)
library(raster)
library(caret)
library(dplyr)
library(tidyr)
library(RStoolbox)
library(ggplot2) 
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling
```

I used the `getLandsat_data()` function in `getSpatialData` package to download Landsat Level-1 imagery from [AWS](  https://aws.amazon.com/what-is-cloud-computing/). However, in order to apply the trained model in 2018 to other years, the surface referance images are required, which should be applied from [USGS earthexplorer](https://earthexplorer.usgs.gov/). So, at last, I download all the data from [USGS earthexplorer](https://earthexplorer.usgs.gov/)

The following code explains how to download Landsat date using `getSpatialData` package.

```{r, eval = FALSE} 
## set aoi and time range for the query
set_aoi(buffaloBoundary_union_cast[1,]$geometry)
time_range <-  c("2018-08-23", "2018-08-23")

# Login to USGS ERS
login_USGS(username = "*****", password = "*****")
set_archive(dataDirectory)

# query for records for AOI, time range and product
query <- getLandsat_query(time_range = time_range)

# preview a record
getLandsat_preview(query[1,])

# download record 1 with level "l1" (will direct to AWS automaticaly)
files <- getLandsat_data(records = query[1,], level = "l1", source = "auto")
tilPath = files[[1]][grepl(".TIF$",files[[1]])]
```

The Landsat 8 imagery in Augest 23, 2018 was used in this study. A fake color image of this image is posted below.

```{r, echo = FALSE, fig.width=5, fig.height=5, fig.cap="Fig1.Landsat 8 image in Augest 23, 2018"}
dataDirectory <- 'LandsatImage'
#get the boundary of Buffalo
buffaloBoundary <- shapefile('LandsatImage/BuffaloBoundary/BuffaloBoundary.shp')
buffaloBoundary = st_as_sf(buffaloBoundary)

#get the training and testing data
labeledROI <- shapefile('LandsatImage/Training data/Training_data.shp')

##get the Landsat data SR
r1 <- raster("LandsatImage/HistoricalImage/LC08_L1TP_017030_20180823_20180829_01_T1_sr_band1.tif")
r2 <- raster("LandsatImage/HistoricalImage/LC08_L1TP_017030_20180823_20180829_01_T1_sr_band2.tif")
r3 <- raster("LandsatImage/HistoricalImage/LC08_L1TP_017030_20180823_20180829_01_T1_sr_band3.tif")
r4 <- raster("LandsatImage/HistoricalImage/LC08_L1TP_017030_20180823_20180829_01_T1_sr_band4.tif")
r5 <- raster("LandsatImage/HistoricalImage/LC08_L1TP_017030_20180823_20180829_01_T1_sr_band5.tif")
r6 <- raster("LandsatImage/HistoricalImage/LC08_L1TP_017030_20180823_20180829_01_T1_sr_band6.tif")
r7 <- raster("LandsatImage/HistoricalImage/LC08_L1TP_017030_20180823_20180829_01_T1_sr_band7.tif")
landsetImage <- brick(r1, r2, r3, r4, r5, r6, r7)
names(landsetImage) <- paste0("B", c(1:7))

plotRGB(landsetImage, r=5, g=4, b=3, stretch = 'lin')

# transform CRS
landset_crs <- raster::crs(landsetImage, asText = TRUE)
buffaloBoundary_transe <- st_transform(buffaloBoundary, crs = landset_crs)
buffaloBoundary_Spatial <- as_Spatial(buffaloBoundary_transe$geom)

# crip Buffalo
BuffaloLandsat <- mask(landsetImage, buffaloBoundary_Spatial) %>% raster::crop(bbox(buffaloBoundary_Spatial))
```
```{r, echo = FALSE, results = 'hide', message=F, warning=F}
rm(r1,r2,r3,r4,r5,r6,r7)
rm(buffaloBoundary_transe)
gc()
```

In order to train the three methods in this study, labeled sample data are required. They were collected in ENVI use visual interpretation. The class of each region was labeled. Then, use the code below to get the spectral information in the Landsat image for each labeled point.

```{r, results = 'hide', message=F, warning=F}
responseCol <- "CLASS_ID"

labeledPoints = data.frame(matrix(vector(), nrow = 0, ncol = length(names(landsetImage)) + 1))   
for (i in 1:length(unique(labeledROI[[responseCol]]))){
  category <- unique(labeledROI[[responseCol]])[i]
  categorymap <- labeledROI[labeledROI[[responseCol]] == category,]
  dataSet <- raster::extract(landsetImage, categorymap)
  if(is(labeledROI, "SpatialPointsDataFrame")){
    dataSet <- cbind(dataSet, class = as.numeric(rep(category, nrow(dataSet))))
    labeledPoints <- rbind(labeledPoints, dataSet[complete.cases(dataSet),])
  }
  if(is(labeledROI, "SpatialPolygonsDataFrame")){
    dataSet <- dataSet[!unlist(lapply(dataSet, is.null))]
    dataSet <- lapply(dataSet, function(x){cbind(x, class = as.numeric(rep(category, nrow(x))))})
    df <- do.call("rbind", dataSet)
    labeledPoints <- rbind(labeledPoints, df)
  }
}
```

After the arrangement of sample data, the sample data were divided into two groups- 80% for training data and 20% for testing data. Then, the training data were used to training the three models: Random Forest, Support Vector Machine and Maximum likelihood.

```{r, results = 'hide', message=F, warning=F}
trainList<-createDataPartition(y=labeledPoints$class,p = 0.8,list=FALSE)
training<-labeledPoints[trainList,]
testing<-labeledPoints[-trainList,]
```
```{r, echo = FALSE, results = 'hide', message=F, warning=F}
rm(labeledROI, labeledPoints)

#rm(buffaloBoundary)
rm(landsetImage)
gc()
```
```{r, results = 'hide', message=F, warning=F}
modFit_rf <- train(as.factor(class) ~ ., method = 'rf', data = training) #random forest
modFit_svm <- train(as.factor(class) ~ ., method = 'svmLinear', data = training) #svmLinear
model_mlc <- train(as.factor(class) ~ ., model = 'mlc', data = training) #Maxmum likelihood
```
```{r,echo = FALSE, results = 'hide', message=F, warning=F}
pre_rf <- predict(modFit_rf, testing)
pre_svm <- predict(modFit_svm, testing)
pre_mlc <- predict(model_mlc, testing)

test_set <- data.frame(obs = factor(testing$class))
test_set$obsClass <- factor(recode(factor(testing$class), "1" = "Vegetation","2" = "Impervious surface ","3" = "Soil","4" = "Water"))
test_set$preRF <- factor(recode(pre_rf, "1" = "Vegetation","2" = "Impervious surface ","3" = "Soil","4" = "Water"))
test_set$preSVM <- factor(recode(pre_svm, "1" = "Vegetation","2" = "Impervious surface ","3" = "Soil","4" = "Water"))
test_set$preMLC <- factor(recode(pre_mlc, "1" = "Vegetation","2" = "Impervious surface ","3" = "Soil","4" = "Water"))

CMrf = confusionMatrix(data = test_set$preRF, reference = test_set$obsClass)
CMsvm = confusionMatrix(data = test_set$preSVM, reference = test_set$obsClass)
CMmlc = confusionMatrix(data = test_set$preMLC, reference = test_set$obsClass)
```
```{r, echo = FALSE, results = 'hide', message=F, warning=F}
rm(modFit_svm, model_mlc, pre_rf, pre_svm, pre_mlc, test_set)
gc()
```

After the training and testing of three models, the best one can be selected. The Landsat image in 2013,2015, 2017 and 2018 in Buffalo could be classified. Those Landsat images were clipped by the boundary of Buffalo city. The clipped 2018 image is as below:

```{r, echo = FALSE, results = 'hide', message=F, warning=F,fig.width=3, fig.height=3, fig.cap="Fig2.Landsat 8 image at Buffalo in 2018"}
plotRGB(BuffaloLandsat, r=5,g=4,b=3, stretch ='lin')
#rf <- writeRaster(preds_rf2018, filename="LandsatImage/Classfied/2018.tif", format="GTiff", overwrite=TRUE)
# classify use random forest
#preds_rf2018 <- predict(modFit_rf, BuffaloLandsat)
beginCluster()
preds_rf2018 <- clusterR(BuffaloLandsat, raster::predict, args = list(model = modFit_rf))
endCluster()

rm(BuffaloLandsat)
gc()
```
After the clipping, Buffalo images were classified using the selected method. Then, the area of each land cover class was calculated to track the change of Buffalo.

```{r, echo = FALSE, results = 'hide', message=F, warning=F}
landsatNames = c("LC08_L1TP_017030_20131012_20170308_01_T1_sr", "LC08_L1TP_017030_20150916_20170225_01_T1_sr", 
                 "LC08_L1TP_017030_20170921_20171012_01_T1_sr")
                 #"LC08_L1TP_017030_20180823_20180829_01_T1_sr")

classifiedImage = stack()
beginCluster()

for(i in 1:length(landsatNames)){
  layername = landsatNames[i]
  path1 <- paste("LandsatImage/HistoricalImage/",layername, "_band", 1, ".tif",sep = "", collapse = "")
  path2 <- paste("LandsatImage/HistoricalImage/",layername, "_band", 2, ".tif",sep = "", collapse = "")
  path3 <- paste("LandsatImage/HistoricalImage/",layername, "_band", 3, ".tif",sep = "", collapse = "")
  path4 <- paste("LandsatImage/HistoricalImage/",layername, "_band", 4, ".tif",sep = "", collapse = "")
  path5 <- paste("LandsatImage/HistoricalImage/",layername, "_band", 5, ".tif",sep = "", collapse = "")
  path6 <- paste("LandsatImage/HistoricalImage/",layername, "_band", 6, ".tif",sep = "", collapse = "")
  path7 <- paste("LandsatImage/HistoricalImage/",layername, "_band", 7, ".tif",sep = "", collapse = "")
  r1 <- raster(path1)
  r2 <- raster(path2)
  r3 <- raster(path3)
  r4 <- raster(path4)
  r5 <- raster(path5)
  r6 <- raster(path6)
  r7 <- raster(path7)
  landsetImageBrick <- brick(r1, r2, r3, r4, r5, r6, r7)
  names(landsetImageBrick) <- paste0("B", c(1:7))

  rm(r1,r2,r3,r4,r5,r6,r7)
  gc()
 
  BuffaloLandsatBrick <- raster::crop(landsetImageBrick,bbox(buffaloBoundary_Spatial))%>% mask(buffaloBoundary_Spatial) 
  
  preds_rf <- clusterR(BuffaloLandsatBrick, raster::predict, args = list(model = modFit_rf))
  
  classifiedImage <- stack(classifiedImage, preds_rf)
  names(classifiedImage)[i] <- paste("C",2011+i*2,sep = "", collapse = "")
 # writePath = paste("LandsatImage/Classfied/",2011+i*2, ".tif",sep = "", collapse = "")
#  rf <- writeRaster(preds_rf, filename=writePath, format="GTiff", overwrite=TRUE)
  
  rm(landsetImageBrick)
  rm(BuffaloLandsatBrick)
  gc()
 
}

endCluster()

gc()
```


# Results

Random Forest, Support Vector Machine and Maximum Likelihood methods were trained and tested in this study. The confusion matrix of those methods were listed below.


```{r,echo = FALSE,  message=F, warning=F}
"Random forest"
CMrf$table
CMrf$overall

```
```{r,echo = FALSE,  message=F, warning=F}
"Support Vector Machine"
CMsvm$table
CMsvm$overall
```
```{r,echo = FALSE,  message=F, warning=F}
"Maximum likelihood"
CMmlc$table
CMmlc$overall

```

According to the confusion matrix and the overall accuracy, Random Forest was selected as a good method for landcover mapping for Buffalo.
The classified map of Buffalo in 2013,2015,2017 and 2018 were posted in Fig.3.

```{r,echo = FALSE, results = 'hide', message=F, warning=F,fig.width=4, fig.height=4}
#Plot map
factorMap2013 <- as.factor(classifiedImage$C2013)
level <- levels(factorMap2013)[[1]]
level[["landcover"]] <- c("Vegetation", "ImperviousSurface ", "Soil", "Water")
levels(factorMap2013) <- level
levelplot(factorMap2013, main="2013 Buffalo", col.regions=rev(terrain.colors(4)), xlab="", ylab="", axes=FALSE)
```
```{r,echo = FALSE, results = 'hide', message=F, warning=F,fig.width=4, fig.height=4}
factorMap2015 <- as.factor(classifiedImage$C2015)
levels(factorMap2015) <- level
levelplot(factorMap2015, main="2015 Buffalo", col.regions=rev(terrain.colors(4)), xlab="", ylab="", axes=FALSE)
```
```{r,echo = FALSE, results = 'hide', message=F, warning=F,fig.width=4, fig.height=4 }
factorMap2017 <- as.factor(classifiedImage$C2017)
levels(factorMap2017) <- level
levelplot(factorMap2017, main="2017 Buffalo", col.regions=rev(terrain.colors(4)), xlab="", ylab="", axes=FALSE)
```
```{r,echo = FALSE, results = 'hide', message=F, warning=F,fig.width=4, fig.height=4, fig.cap="Fig3. Classified image at Buffalo"}
factorMap2018 <- as.factor(preds_rf2018)
levels(factorMap2018) <- level
levelplot(factorMap2018, main="2018 Buffalo", col.regions=rev(terrain.colors(4)), xlab="", ylab="", axes=FALSE)

```

According to the classfied maps, the change of every landcover class can be derived.

```{r,echo = FALSE, results = 'hide', message=F, warning=F,fig.width=6, fig.height=3, fig.cap="Fig4.Land cover change of Buffalo"}
####plot the change of landcover landuse
Class2013 <- raster::extract(classifiedImage$C2013, buffaloBoundary_Spatial)
Class2015 <- raster::extract(classifiedImage$C2015, buffaloBoundary_Spatial)
Class2017 <- raster::extract(classifiedImage$C2017, buffaloBoundary_Spatial)
Class2018 <- raster::extract(preds_rf2018, buffaloBoundary_Spatial)


data = data.frame( year = c(2013,2015,2017,2018),
                   Vegetation = c(sum(Class2013[[1]] == 1),sum(Class2015[[1]] == 1),sum(Class2017[[1]] == 1),sum(Class2018[[1]] == 1)),
                   ImperviousSurface = c(sum(Class2013[[1]] == 2),sum(Class2015[[1]] == 2),sum(Class2017[[1]] == 2),sum(Class2018[[1]] == 2)),
                   Soil = c(sum(Class2013[[1]] == 3),sum(Class2015[[1]] == 3),sum(Class2017[[1]] == 3),sum(Class2018[[1]] == 3)),
                   Water = c(sum(Class2013[[1]] == 4),sum(Class2015[[1]] == 4),sum(Class2017[[1]] == 5),sum(Class2017[[1]] == 4)))

dataWrangle = gather(data, "class", "area/km2", 2:5)
dataWrangle$`area/km2` = dataWrangle$`area/km2`*0.9/1000
ggplot(dataWrangle, aes(x=year, y=`area/km2`)) + 
  geom_line(aes(colour = factor(class)))

```


# Conclusions

This study test three classification method in mapping the land cover in Buffalo. It showed that Random Forest performed good when doing the coarse classisication. However, the change inside every land cover can not be detected with this method. Thus, more comprehensive methods are under requirement for the detail change detection.

# References
[1] Mountrakis, G., Im, J., & Ogole, C. (2011). Support vector machines in remote sensing: A review. ISPRS Journal of Photogrammetry and Remote Sensing, 66(3), 247-259.

[2] Pal, M. (2005). Random forest classifier for remote sensing classification. International Journal of Remote Sensing, 26(1), 217-222.

[3]Strahler, A. H. (1980). The use of prior probabilities in maximum likelihood classification of remotely sensed data. Remote sensing of Environment, 10(2), 135-163.
